---
title: "Lab5"
author: "Shuchen Wu"
format: html
editor: visual
embed-resources: true
fig-width: 8
fig-height: 6
---

## Step 1

```{r}
# install.packages("data.table")
library(data.table)
```

## Step 2

```{r}
url <- 'https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/covid/population.csv'
if (!file.exists("covid_processed.csv"))
  download.file(
    url = url,
    destfile = "population.csv",
    method   = "libcurl",
    timeout  = 60
    )
population <- read.csv('population.csv')
```

## Step 3

```{r}
# install.packages("janitor")
library(tidyverse)
library(janitor)
population <- row_to_names(population, 1)
population <- population %>%
  select(-1, -state)
population <- population %>%
  rename(state_name = NAME)

abb_map <- setNames(state.abb, state.name)
abb_map["District of Columbia"] <- "DC"
abb_map["Puerto Rico"] <- "PR"

population <- population %>%
  mutate(state = abb_map[state_name])

population <- population %>%
  pivot_longer(
    cols = starts_with("POP"),
    names_to = "year",
    values_to = "population"
  )

head(population)
```

## Step 4

```{r}
library(jsonlite)
url <- "https://github.com/datasciencelabs/2024/raw/refs/heads/main/data/regions.json"
regions <- fromJSON(url)
regions <- regions %>%
  mutate(region = unlist(region))
regions <- regions %>%
  unnest(states) %>%
  transmute(
    state_name = states,
    region = factor(region),
    region_name = region_name
  )
regions <- regions %>%
  mutate(
    region_name = if_else(region == 2,
                          "NY & NJ + PR + VI",
                          region_name)
  )

head(regions, 10)
```

## Step 5

```{r}
population <- population %>%
  left_join(regions, by = "state_name")

head(population)
```

## Step 6

```{r}
library(httr2)
api <- "https://data.cdc.gov/resource/pwn4-m3yp.json"
request <- request(paste0(api,paste0("?$limit=10000000000")))
response <- request |> req_perform() |> resp_body_string()
cases_raw <- fromJSON(response)
head(cases_raw)
```

If we replace "?\$limit=10000000000" with an empty string, the API will return only the default number of rows instead of the full dataset. Most Socrata/CDC APIs apply a default limit (often around 1000 rows) when \$limit is not specified. As a result, cases_raw would contain only a partial subset of the data, which could lead to incorrect summaries or analyses because many records would be missing.

## Step 7

```{r}
library(lubridate)
cases <- cases_raw %>%
  transmute(
    state = state,
    date  = as_date(ymd_hms(end_date)),
    cases = as.numeric(new_cases)
  )

head(cases)
```

## Step 8

```{r}
data_full <- cases %>%
  left_join(population, by = "state")

data_full <- data_full %>%
  mutate(
    epi_week = epiweek(date),
    epi_year = epiyear(date)
  )

data_full <- data_full %>%
  arrange(state, date)

head(data_full)

```
